From Net-Zero to Net-Negative: Assessing \hydrogen{} and \carbon{} Network Strategies in Europe

### 2023-11-23 report/submissions/journal/paper/paper.tex
add subsections
review and add to subsections 1 and 2 in Results

@@

Hydrogen can be gained from electrolysis and steam methane reforming (SMR). Geological distribution and potentials for \hydrogen storage in salt caverns are derived from~\cite{caglayanTechnicalPotentialSalt2020}. Re-electrification of hydrogen is possible via fuel cells. If enabled, hydrogen can be transported via pipelines between regions which can be expanded without limit. The topology of the \hydrogen network is mirroring the topology of the electricity network.

@@

\begin{figure*}[h]
    \centering
    \begin{subfigure}{\linewidth}
        \centering
        \includegraphics[width=0.95\linewidth]{co2-only/figures/90_nodes/operation_map_carbon.png}
        \caption{\carbon{} Sector.}
        \label{fig:operation_map_carbon}
    \end{subfigure}
    \begin{subfigure}{\linewidth}
        \centering
        \includegraphics[width=0.95\linewidth]{h2-only/figures/90_nodes/operation_map_hydrogen.png}
        \caption{\hydrogen{} Sector.}
        \label{fig:operation_map_hydrogen}
    \end{subfigure}
    \caption{Optimal operation per sector for a net-zero energy system in Europe with average production on the left and average consumption on the right for both, (a) the \carbon sector in the \carbon{}-Grid model and (b) the \hydrogen{} sector in the \hydrogen-Grid model.}
    \label{fig:operation_map}
\end{figure*}

\subsection*{}

Examining the cost-reductions across the different models raises an important question: why is the hydrogen network more cost-effective than its carbon counterpart, and how does it achieve similar cost reductions as the hybrid model? To answer this question, it is crucial to understand the underlying mechanisms and geographical implications of the two transportation systems. Fig.~\ref{fig:operation_map_carbon} maps the average production, consumption and flow in the carbon sector for the \carbongrid model. Fig. \ref{fig:operation_map_hydrogen} shows equivalent quantities in the hydrogen sector for the \hydrogengrid model. In each region, circles divide into contributions from technologies, and their sizes correspond proportionally to production on the left and consumption on the right. The region's colors represent the average price of the considered carrier.
The two fundamental differences between the grid strategies are quickly evident: The \carbongrid model captures large amounts of carbon from bioenergy and industrial processes, especially in central and eastern Europe, at low cost. The carbon grid transports cheap carbon to regions where it is sequestered in offshore caverns or processed into synfuel, methane and methanol together with cheaply available green hydrogen. End-use primarily depends on the quality of renewable resources. In Italy, Greece, North Germany, and Denmark, the transported carbon is predominantly sequestered and reused to some extent. Conversely, in Spain and the United Kingdom, which both have high renewable potentials, the transported carbon is fully utilized, complementing air-captured carbon. Thus, the implementation of a \carbon network allows for connecting regions with abundant, low-price carbon to regions with carbon scarcity where in the absence of a \carbon network DAC facilities had to provide most carbon.
% Sequestered carbon is nearly always transported before

The \hydrogengrid on the other hand, transports cheap hydrogen from regions with abundant renewable resources to regions with high energy demand and abundant emissions. The main transportation routes go from Spain and the United Kingdom to Central Europe. In these regions, industries use the hydrogen to fuel processes and land transport. Additionally, the imported hydrogen allows for the direct reuse of local \carbon emissions in synfuel production, leading to decentralized synfuel production across the continent. Hydrogen produced in Italy contributes to local methane production, which then travels to Central Europe via gas pipelines to supply CHPs and industries."

% Therefore, one can conclude that a \carbon network favors a decentralized CC and a CU centralized at regions with abundant renewable resources. A \hydrogen network favors a centralized electrolysis system at regions with abundant renewable resources and a decentralized CU system.
These findings highlight the pivotal advantage of the \hydrogen network in contrast to the \carbon network. Central Europe with its industrial clusters and dense population is characterized by a scarcity in energy and a surplus and abundance of \carbon. While the \carbon network cost-efficiently resolves the surplus of carbon, the hydrogen resolves both the energy scarcity and the carbon surplus.

@@

As depicted in Fig.~\ref{fig:capacity_maps}, the hybrid model incorporates the advantages from the a large hydrogen network and complements it with a carbon network. Mirroring the behavior of \hydrogengrid model, it transports hydrogen from Western regions like Spain and the UK to Central Europe, with a centralized electrolysis and decentralized carbon utilization. The carbon network complements this by transporting cheap carbon to sequestration sites in the North Sea, Baltic Sea, Greece, and Italy, minimizing the need for Direct Air Capture (DAC) at these locations. A notable aspect of the hybrid model is the lack of overlap in the routes used by the hydrogen and carbon networks. The advantages of this model include a broader range of technologies contributing to its robustness, reduced reliance on DAC (which may be costlier than anticipated), and less land use for wind and solar due to the decreased necessity for DAC.

@@

We now examine the case of moving from a net zero emissions target to a net emissions reduction target. Fig.~\ref{fig:net-negative_cost_bar} illustrates the shift in investments for the \hydrogengrid and the Hybrid model between the net-zero and the net-negative scenario.

The \hydrogengrid model emphasizes a substantial investment in hydrogen infrastructure, signifying a strategic bet on hydrogen's role in decarbonization. There is also a significant allocation for methanation and the expansion of electricity grids, underscoring the integration of hydrogen production with renewable energy sources.

>>>>

\section{Introduction}

@@

Hydrogen can be gained from electrolysis and steam methane reforming (SMR). Geological distribution and potentials for \hydrogen{} storage in salt caverns are derived from~\cite{caglayanTechnicalPotentialSalt2020}. Re-electrification of hydrogen is possible via fuel cells. If enabled, hydrogen can be transported via pipelines between regions which can be expanded without limit. The topology of the \hydrogen{} network is mirroring the topology of the electricity network.

@@

\begin{figure*}[h]
    \centering
    \begin{subfigure}{\linewidth}
        \centering
        \includegraphics[width=0.95\linewidth]{co2-only/figures/90_nodes/operation_map_carbon.png}
        \caption{\carbon{} Sector.}
        \label{fig:operation_map_carbon}
    \end{subfigure}
    \begin{subfigure}{\linewidth}
        \centering
        \includegraphics[width=0.95\linewidth]{h2-only/figures/90_nodes/operation_map_hydrogen.png}
        \caption{\hydrogen{} Sector.}
        \label{fig:operation_map_hydrogen}
    \end{subfigure}
    \caption{Optimal operation per sector for a net-zero energy system in Europe with average production on the left and average consumption on the right for both, (a) the \carbon{} sector in the \carbon{}-Grid model and (b) the \hydrogen{} sector in the \hydrogen-Grid model.}
    \label{fig:operation_map}
\end{figure*}

\subsection*{Superiority of \hydrogen{} network}

Examining the cost-reductions across the different models raises an important question: why is the hydrogen network more cost-effective than its carbon counterpart, and how does it achieve similar cost reductions as the hybrid model? To answer this question, it is crucial to understand the underlying mechanisms and geographical implications of the two transportation systems. Fig.~\ref{fig:operation_map_carbon} maps the average production, consumption and flow in the carbon sector for the \carbongrid{} model. Fig. \ref{fig:operation_map_hydrogen} shows equivalent quantities in the hydrogen sector for the \hydrogengrid{} model. In each region, circles divide into contributions from technologies, and their sizes correspond proportionally to production on the left and consumption on the right. The region's colors represent the average price of the considered carrier.

The two fundamental differences between the grid strategies are quickly evident: The \carbongrid{} model captures large amounts of carbon from bioenergy and industrial processes, especially in central and eastern Europe, at low cost. The carbon grid transports cheap carbon to regions where it is sequestered in offshore caverns or processed into synfuel, methane and methanol together with cheaply available green hydrogen. End-use primarily depends on the quality of renewable resources, and thus the availability of cheap hydrogen. In locations with relatively poor renewables and thus higher costs of hydrogen production, such as Italy, Greece, Northern Germany, and lcoations around the Baltic Sea, the transported carbon is predominantly sequestered. Conversely, in Spain and the United Kingdom, which both have high renewable potentials and cheap hydrogen production costs, the transported carbon is combined with hydrogen and fed into CU processes, with additional carbon being captured via DAC also running on the great renewable resources.

In summary, the implementation of a \carbon{} network allows for connecting regions with abundant, low-price carbon (from point sources) to regions that can sequester the carbon or that can produce cheap hydrogen to make CU products (FT fuel, methanol, synthetic methane). In the absence of a \carbon{} network, sequestration sites first sequester all carbon from cheap point sources, and deploy additional DAC facilities at the locations with cheap renewables to capture the remaining necessary carbon for sequestration. Then, synfuel demand is met at the locations which offer the best mix of cheap carbon from point sources and cheap hydrogen from local renewables.
% Sequestered carbon is nearly always transported before

The \hydrogengrid{} on the other hand, transports cheap hydrogen from regions with abundant renewable resources to regions with high energy demand and abundant emissions. The main transportation routes go from Spain and the United Kingdom to Central Europe. In these regions, industries use the hydrogen for demand in industry and fuel cell land transport, and to produce synfuels with locally captured carbon from point sources. This leads to a very decentralized synfuel production across the continent. Furthermore, the introduction of a hydrogen network enables to reduce both DAC production and FT fuels on the British Isles and Iberian Peninsula. In Great Britain an Ireland, the remaining local carbon is then sequestered, and all the hydrogen that previously went into FT fuel production is exported to Central Europe (see above). In Spain and Portugal, some DAC remains and the carbon sourced from DAC is used, along with locally captured carbon from point sources, to produce FT fuels (but less than without a hydrogen network). Surplus hydrogen from lower FT fuel production is exported to Central Europe.

% Therefore, one can conclude that a \carbon{} network favors a decentralized CC and a CU centralized at regions with abundant renewable resources. A \hydrogen{} network favors a centralized electrolysis system at regions with abundant renewable resources and a decentralized CU system.
These findings explain the advantage of the \hydrogen{} network in contrast to the \carbon{} network. Central Europe with its industrial clusters and dense population is characterized by a scarcity in hydrogen and a surplus and abundance of \carbon. While the \carbon{} network cost-efficiently resolves the surplus of carbon in Central Europe by exporting it to regions with cheap renewable resources to produce synfuels, it needs to supply direct hydrogen demand in Central Europe (for industry and fuel cell land transport) with expensive local hydrogen production. The hydrogen network resolves both the energy scarcity by supplying hydrogen demand in Central Europe with cheap hydrogen imported from Spain and Great Britain, and the carbon surplus by transporting hydrogen to the sites with cheaper carbon available in Central Europe, and producing FT fuels highly decentralized throughout all of Central Europe.

@@

\subsection*{Flexibility gains through hybrid configuration}

As depicted in Fig.~\ref{fig:capacity_maps}, the hybrid model incorporates the advantages from the a large hydrogen network and complements it with a carbon network. The carbon network unlocks cheap carbon in landlocked locations like  Poland, Bulgaria, and Austria and transports it to sequestration sites in the North Sea, Baltic Sea, Greece, and Italy, minimizing the need for DAC directly at the sequestration sites. At the same time, there is reduced need to transport hydrogen from Western regions like Spain and the UK to Central Europe.

A notable aspect of the hybrid model is the lack of overlap in the routes used by the hydrogen and carbon networks. The advantages of this model include a broader range of technologies contributing to its robustness, reduced reliance on DAC (which may be costlier than anticipated), and less land use for wind and solar due to the decreased necessity for DAC.

In summary, sequestration moves to sites with or close to the cheapest carbon sources. Adding the carbon network unlocks cheap inland carbon point sources to be transported to sequestration. FT fuel production moves to sites with or close to cheap hydrogen. The model prefers sites with locally available carbon from point sources and transport hydrogen from nearby cheap hydrogen producing locations, rather than transporting the carbon to the hydrogen site.

Furthermore, there are qualitative advantages to building out both a hydrogen and a carbon network that are not captured in the strict cost-benefit calculation employed in our optimization model. The resulting system may be more resilient towards unknown future developments in technologies, geopolitics, or policy goals.

@@

We now examine the case of moving from a net zero emissions target to a net emissions reduction target. Fig.~\ref{fig:net-negative_cost_bar} illustrates the shift in investments for the \hydrogengrid{} and the Hybrid model between the net-zero and the net-negative scenario.

The \hydrogengrid{} model emphasizes a substantial investment in hydrogen infrastructure, signifying a strategic bet on hydrogen's role in decarbonization. There is also a significant allocation for methanation and the expansion of electricity grids, underscoring the integration of hydrogen production with renewable energy sources.
